---
title: "01 Load and QC"
output: html_document
---

```{r setup, include=FALSE}
# If running interactively from within /scripts, move up to project root
if (basename(getwd()) == "scripts") setwd("..")

# Now we are in the project root
knitr::opts_knit$set(root.dir = normalizePath(".")) 
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)


source("scripts/00_setup.R")
```

Load data:

```{r load-10x-matrices}
lib_ids <- libraries$library_id
tenx_list <- setNames(vector("list", length(lib_ids)), lib_ids)

for (id in lib_ids) {
  msg("Reading 10x data for ", id)
  tenx_list[[id]] <- read_10x_for_library(id)
}
```

```{r}
prol.data <- tenx_list$Lib1
diff_ctl.data <- tenx_list$Lib2
diff_low_TNF.data <- tenx_list$Lib3
diff_high_TNF.data<- tenx_list$Lib4
```

```{r}
diff_ctl_rna <- diff_ctl.data$`Gene Expression`
diff_ctl_hto <- diff_ctl.data$`Antibody Capture`

diff_low_TNF_rna <- diff_low_TNF.data$`Gene Expression`
diff_low_TNF_hto <- diff_low_TNF.data$`Antibody Capture`

diff_high_TNF_rna <- diff_high_TNF.data$`Gene Expression`
diff_high_TNF_hto <- diff_high_TNF.data$`Antibody Capture`

prol_rna <- prol.data$`Gene Expression`
prol_hto <- prol.data$`Antibody Capture`

# Remove the first row (only differentiation)
diff_ctl_hto <- diff_ctl_hto[-c(1),]
diff_low_TNF_hto <- diff_low_TNF_hto[-c(1),]
diff_high_TNF_hto <- diff_high_TNF_hto[-c(1),]

# Find overlapping cell barcodes 
join_diff_ctl <- intersect(colnames(diff_ctl_rna), colnames(diff_ctl_hto))
join_diff_low <- intersect(colnames(diff_low_TNF_rna), colnames(diff_low_TNF_hto))
join_diff_high <- intersect(colnames(diff_high_TNF_rna), colnames(diff_high_TNF_hto))
joint_prol <- intersect(colnames(prol_rna), colnames(prol_hto))

# Subset RNA and HTO counts by joint cell barcodes 
diff_ctl_rna <- diff_ctl_rna[, join_diff_ctl]
diff_ctl_hto <- as.matrix(diff_ctl_hto[, join_diff_ctl])

diff_low_TNF_rna <- diff_low_TNF_rna[, join_diff_low]
diff_low_TNF_hto <- as.matrix(diff_low_TNF_hto[, join_diff_low])

diff_high_TNF_rna <- diff_high_TNF_rna[, join_diff_high]
diff_high_TNF_hto <- as.matrix(diff_high_TNF_hto[, join_diff_high])

prol_rna <- prol_rna[, joint_prol]
prol_hto <- as.matrix(prol_hto[, joint_prol])

# Confirm that the HTO have the correct names. Should be TotalSeqB1,2,and3
rownames(diff_ctl_hto)
rownames(diff_low_TNF_hto)
rownames(diff_high_TNF_hto)
rownames(prol_hto)
```
Create Seurat objects 

```{r}
# Setup Seurat object 
dif_ctl_seurat <- CreateSeuratObject(counts = diff_ctl_rna, project = "Diff_control")
dif_low_TNF_seurat <- CreateSeuratObject(counts = diff_low_TNF_rna, project = "Diff_low")
dif_high_TNF_seurat <- CreateSeuratObject(counts = diff_high_TNF_rna, project = "Diff_high")
prol_seurat <- CreateSeuratObject(counts = prol_rna, project = "Proliferation")
```

Adding, normalising, and demultiplexing HTO data

```{r}
# Create a new assay to store the HTO information and add this to the Seurat object
dif_ctl_seurat[["HTO"]] <- CreateAssayObject(counts = diff_ctl_hto)
dif_low_TNF_seurat[["HTO"]] <- CreateAssayObject(counts = diff_low_TNF_hto)
dif_high_TNF_seurat[["HTO"]] <- CreateAssayObject(counts = diff_high_TNF_hto)
prol_seurat[["HTO"]] <- CreateAssayObject(counts = prol_hto)
```

We can check that we now two assays - RNA and HTO

```{r}
# Validate that we now have two assays 
Assays(dif_ctl_seurat)
Assays(dif_low_TNF_seurat)
Assays(dif_high_TNF_seurat)
Assays(prol_seurat)
```

We can check the default assay

```{r}
# Check defualt assay
DefaultAssay(dif_ctl_seurat)
DefaultAssay(dif_low_TNF_seurat)
DefaultAssay(dif_high_TNF_seurat)
DefaultAssay(prol_seurat)
```

Now we normalise the HTO data

```{r}
# Normalize HTO data, here we use centered log-ratio (CLR) transformation
dif_ctl_seurat <- NormalizeData(dif_ctl_seurat, assay = "HTO", normalization.method = "CLR")
dif_low_TNF_seurat <- NormalizeData(dif_low_TNF_seurat, assay = "HTO", normalization.method = "CLR")
dif_high_TNF_seurat <- NormalizeData(dif_high_TNF_seurat, assay = "HTO", normalization.method = "CLR")
prol_seurat <- NormalizeData(prol_seurat, assay = "HTO", normalization.method = "CLR")
```

And now demultipex the data:

```{r}
# Demultiplex cells based on HTO enrichment
dif_ctl_seurat <- HTODemux(dif_ctl_seurat, assay = "HTO", positive.quantile = 0.99)
dif_low_TNF_seurat <- HTODemux(dif_low_TNF_seurat, assay = "HTO", positive.quantile = 0.99)
dif_high_TNF_seurat <- HTODemux(dif_high_TNF_seurat, assay = "HTO", positive.quantile = 0.99)
prol_seurat <- HTODemux(prol_seurat, assay = "HTO", positive.quantile = 0.99)
```

We can visualize how many cells are classified as singlets, doublets and negative/ambiguous cells.

```{r}
# Global classification results
table(dif_ctl_seurat$HTO_classification.global)
```

```{r}
# Global classification results
table(dif_low_TNF_seurat$HTO_classification.global)
```

```{r}
# Global classification results
table(dif_high_TNF_seurat$HTO_classification.global)
```

```{r}
# Global classification results
table(prol_seurat$HTO_classification.global)
```

Now merge the differentiation samples samples: 

```{r}
combined <- merge(dif_ctl_seurat, y = c(dif_low_TNF_seurat, dif_high_TNF_seurat), add.cell.ids = c("Control", "Low", 'High'), project = "TN")
```

Now we extract the singlets:

```{r}
Idents(combined) <- "HTO_classification.global"
Idents(prol_seurat) <- "HTO_classification.global"

# Extract the singlets
combined <- subset(combined, idents = "Singlet")
prol_seurat <- subset(prol_seurat, idents = "Singlet")
```

Adding easy to read labels (differentiation)

```{r}
metadata_combined <- combined@meta.data %>% 
  mutate(sample.ID = case_when(
            orig.ident == "Diff_control" & hash.ID == "HTO-AHH02-TotalSeqB" ~ "Differentiation control (7 days)",
            orig.ident == "Diff_control" & hash.ID == "HTO-AHH03-TotalSeqB" ~ "Differentiation control (14 days)",
            orig.ident == "Diff_low" & hash.ID == "HTO-AHH02-TotalSeqB" ~ "Differentiation low TNF (7 days)",
            orig.ident == "Diff_low" & hash.ID == "HTO-AHH03-TotalSeqB" ~ "Differentiation low TNF (14 days)",
            orig.ident == "Diff_high" & hash.ID == "HTO-AHH02-TotalSeqB" ~ "Differentiation high TNF (7 days)",
            orig.ident == "Diff_high" & hash.ID == "HTO-AHH03-TotalSeqB" ~ "Differentiation high TNF (14 days)"))

# Add new metadata
combined@meta.data <- metadata_combined
```
Adding easy to read labels (proliferation)
```{r}
metadata_prol_seurat <- prol_seurat@meta.data %>% 
  mutate(sample.ID = case_when(
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH01-TotalSeqB" ~ "Proliferation control",
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH02-TotalSeqB" ~ "Proliferation low TNF",
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH03-TotalSeqB" ~ "Proliferation high TNF"))

# Add new metadata
prol_seurat@meta.data <- metadata_prol_seurat
```

Adding easy to read labels for treatment (differentiation)

```{r}

combined$treatment <- NULL

metadata_combined <- combined@meta.data %>% 
  mutate(treatment = case_when(
            orig.ident == "Diff_control" ~ "Control",
            orig.ident == "Diff_low" ~ "Low TNF",
            orig.ident == "Diff_high" ~ "High TNF"))

# Add new metadata
combined@meta.data <- metadata_combined

```

Adding easy to read labels for treatment (proliferation)

```{r}

prol_seurat$treatment <- NULL

metadata_prol_seurat <- prol_seurat@meta.data %>% 
  mutate(treatment = case_when(
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH01-TotalSeqB" ~ "Control",
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH02-TotalSeqB" ~ "Low TNF",
            orig.ident == "Proliferation" & hash.ID == "HTO-AHH03-TotalSeqB" ~ "High TNF"))

# Add new metadata
prol_seurat@meta.data <- metadata_prol_seurat
```

Calculate novelty score:

```{r}
# Add number of genes per UMI for each cell to metadata
combined$log10GenesPerUMI <- log10(combined$nFeature_RNA) / log10(combined$nCount_RNA)
prol_seurat$log10GenesPerUMI <- log10(prol_seurat$nFeature_RNA) / log10(prol_seurat$nCount_RNA)
```

Perentage of reads that map to the mitochondrial genome

```{r}
combined[["percent.mt"]] <- PercentageFeatureSet(combined, pattern = "^MT-")
prol_seurat[["percent.mt"]] <- PercentageFeatureSet(prol_seurat, pattern = "^MT-")
```

Plot the percentage of mitochondrial reads (differentiation): 
```{r}

combined$treatment <- factor(
  combined$treatment,
  levels = c("Control", "Low TNF", "High TNF")
)

differentiation <- VlnPlot(combined, features = 'percent.mt', group.by = 'treatment' )
```

Plot the percentage of mitochondrial reads (proliferation): 
```{r}
prol_seurat$treatment <- factor(
  prol_seurat$treatment,
  levels = c("Control", "Low TNF", "High TNF")
)

proliferation <- VlnPlot(prol_seurat, features = 'percent.mt', group.by = 'treatment' )
```

Metadata differentiation

```{r}
# Create metadata dataframe
metadata_dif <- combined@meta.data

# Add cell IDs to metadata
metadata_dif$cells <- rownames(metadata_dif)

# Rename columns
metadata_dif <- metadata_dif %>%
        dplyr::rename(seq_folder = orig.ident,
                      nUMI = nCount_RNA,
                      nGene = nFeature_RNA)

# Add metadata back to Seurat object
combined@meta.data <- metadata_dif
```

Metadata proliferation

```{r}
# Create metadata dataframe
metadata_prol <- prol_seurat@meta.data

# Add cell IDs to metadata
metadata_prol$cells <- rownames(metadata_prol)

# Rename columns
metadata_prol <- metadata_prol %>%
        dplyr::rename(seq_folder = orig.ident,
                      nUMI = nCount_RNA,
                      nGene = nFeature_RNA)

# Add metadata back to Seurat object
prol_seurat@meta.data <- metadata_prol
```

Dianostic plots - cell counts per sample


```{r}
# Visualize the number of cell counts per sample - DIF
metadata_dif %>% 
  	ggplot(aes(x=sample.ID, fill=sample.ID)) + 
  	geom_bar() +
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells")
```

```{r}
# Visualize the number of cell counts per sample - Prol
metadata_prol %>% 
  	ggplot(aes(x=sample.ID, fill=sample.ID)) + 
  	geom_bar() +
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells")
```

Transcripts per cell - we can see that majority of our cells in both samples have 1000 UMIs or greater.

```{r}
# Visualize the number UMIs/transcripts per cell
metadata_dif %>% 
  	ggplot(aes(color=sample.ID, x=nUMI, fill= sample.ID)) + 
  	geom_density(alpha = 0.2) + 
    scale_x_continuous(limits = c(0,100000)) + 
    geom_vline(xintercept = 1000)+ 
    geom_vline(xintercept = 45000)
```

```{r}
# Visualize the number UMIs/transcripts per cell
metadata_prol %>% 
  	ggplot(aes(color=sample.ID, x=nUMI, fill= sample.ID)) + 
  	geom_density(alpha = 0.2) + 
    scale_x_continuous(limits = c(0,100000)) + 
    geom_vline(xintercept = 1000)+ 
    geom_vline(xintercept = 45000)
```
Number of genes in each cell:

```{r}
# Visualize the distribution of genes detected per cell via histogram
metadata_dif %>% 
  	ggplot(aes(color=sample.ID, x=nGene, fill= sample.ID)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_continuous(limits = c(0,20000))   +
    geom_vline(xintercept = 800)+ 
    geom_vline(xintercept = 8000)
```

```{r}
# Visualize the distribution of genes detected per cell via histogram
metadata_prol %>% 
  	ggplot(aes(color=sample.ID, x=nGene, fill= sample.ID)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_continuous(limits = c(0,20000))   +
    geom_vline(xintercept = 800)+ 
    geom_vline(xintercept = 8000)
```

Complexity

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI (novelty score)
metadata_dif %>%
  	ggplot(aes(x=log10GenesPerUMI, color = sample.ID, fill=sample.ID)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)
```

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI (novelty score)
metadata_prol %>%
  	ggplot(aes(x=log10GenesPerUMI, color = sample.ID, fill=sample.ID)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)
```

Generally, we expect the novelty score to be above 0.80 for good quality cells. Looks god!

```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
metadata_dif %>% 
  	ggplot(aes(color=sample.ID, x=percent.mt, fill=sample.ID)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	geom_vline(xintercept = 15)
```

```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
metadata_prol %>% 
  	ggplot(aes(color=sample.ID, x=percent.mt, fill=sample.ID)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	geom_vline(xintercept = 15)
```

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
metadata_dif %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~sample.ID)
```

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
metadata_prol %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~sample.ID)
```
Filtering - Number of cells before filtering:
```{r}
ncol(combined)
ncol(prol_seurat)
```

```{r}
# Filter out low quality cells using selected thresholds - these will change with experiment
combined_cell_filtered <- subset(x = combined, 
                         subset= (nUMI > 1000) & (nUMI < 45000) &
                           (nGene > 800) & (nGene < 8000)&
                           (log10GenesPerUMI > 0.80) & 
                           (percent.mt < 15))

prol_cell_filtered <- subset(x = prol_seurat, 
                         subset= (nUMI > 1000) & (nUMI < 45000) &
                           (nGene > 800) & (nGene < 8000)&
                           (log10GenesPerUMI > 0.80) & 
                           (percent.mt < 15))
```

```{r}
ncol(combined_cell_filtered)
ncol(prol_cell_filtered)
```

Gene level filtering:

Within our data we will have many genes with zero counts. These genes can dramatically reduce the average expression for a cell and so we will remove them from our data. We will start by identifying which genes have a zero count in each cell:

```{r} 
##################### DIF
combined_cell_filtered[["RNA"]] <- JoinLayers(combined_cell_filtered[["RNA"]])

# Extract counts
counts <- combined_cell_filtered[["RNA"]]$counts

# Output a logical matrix specifying for each gene on whether or not there are more than zero counts per cell
nonzero_dif <- counts > 0

# Sums all TRUE values and returns TRUE if more than 5 TRUE values per gene
keep_genes_dif <- Matrix::rowSums(nonzero_dif) >= 5

# Only keeping those genes expressed in more than 5 cells
filtered_counts_dif <- counts[keep_genes_dif, ]
```


```{r}
# Extract counts
counts <- prol_cell_filtered[["RNA"]]$counts

# Output a logical matrix specifying for each gene on whether or not there are more than zero counts per cell
nonzero_prol <- counts > 0

# Sums all TRUE values and returns TRUE if more than 5 TRUE values per gene
keep_genes_prol <- Matrix::rowSums(nonzero_prol) >= 5

# Only keeping those genes expressed in more than 5 cells
filtered_counts_prol <- counts[keep_genes_prol, ]
```


Finally, take those filtered counts and create a new Seurat object for downstream analysis.

```{r}
# Reassign to filtered Seurat object
combined_cell_filtered <- CreateSeuratObject(filtered_counts_dif, meta.data = combined_cell_filtered@meta.data )
prol_cell_filtered <- CreateSeuratObject(filtered_counts_prol, meta.data = prol_cell_filtered@meta.data)
```

Number of genes and cells after QC: 
```{r}
ncol(combined_cell_filtered)
ncol(prol_cell_filtered)

nrow(combined_cell_filtered)
nrow(prol_cell_filtered)
```


```{r}
# Save filtered subset to new metadata
metadata_clean_dif <- combined_cell_filtered@meta.data
metadata_clean_prol <- prol_cell_filtered@meta.data
```

Save QC'ed object: 

```{r}
save(combined_cell_filtered, file = "results/qc_diff_merged.RData")
save(prol_cell_filtered, file = "results/prol_cell_filtered.RData")
```

Plot the QC'ed data:

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
metadata_clean_dif %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~sample.ID)
```


```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
metadata_clean_prol %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~sample.ID)
```