---
title: "02 Normalisation and Clustering"
output: html_document
---

```{r setup, include=FALSE}
# If running interactively from within /scripts, move up to project root
if (basename(getwd()) == "scripts") setwd("..")

# Now we are in the project root
knitr::opts_knit$set(root.dir = normalizePath(".")) 
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)


source("scripts/00_setup.R")
```

# Load the quality controled data

```{r}
# load diff
load(file = "results/qc_diff_merged.RData")

# load prol
load(file = "results/prol_cell_filtered.RData")
```

# Add type 

```{r}
combined_cell_filtered$type <-'Differentation'
prol_cell_filtered$type <-'Proliferation'
```

# Merge

```{r}
obj_merged <- merge(combined_cell_filtered, y = prol_cell_filtered, add.cell.ids = c("Diff", "Prol"))
```

# Check

```{r}
class(obj_merged[['RNA']])
Layers(obj_merged[['RNA']])
```
# Normalization and variance stabilization

```{r}
# Confirm: split layers should be gone or empty, and "counts" should exist
Layers(obj_merged[["RNA"]])

DefaultAssay(obj_merged) <- "RNA"   # SCTransform reads RNA counts

obj_merged <- SCTransform(obj_merged, method = "glmGamPoi", verbose = FALSE)

DefaultAssay(obj_merged) <- 'SCT'

obj_merged1 <- obj_merged
```

# Run PCA
```{r}
obj_merged1 <- RunPCA(obj_merged1, dims=1:50)
```

```{r}
obj_merged1 <- FindNeighbors(obj_merged1, dims = 1:50, verbose = FALSE)

obj_merged1 <- FindClusters(obj_merged1, verbose = FALSE,resolution = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1,1.1,1.2))

obj_merged1 <- RunUMAP(obj_merged1, dims = 1:50, verbose = FALSE, seed.use = 42)

obj_merged1 <- PrepSCTFindMarkers(obj_merged1, assay = "SCT", verbose = TRUE)

#all_markers <- FindAllMarkers(obj_merged1,test.use = "wilcox")

#all_markers %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% slice_head(n = 10) %>% ungroup() -> top10
```

```{r}
DimPlot(obj_merged1, split.by = 'sample.ID', ncol = 3)
```

```{r}
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.1", label = TRUE) + ggtitle("res = 0.1")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.2", label = TRUE) + ggtitle("res = 0.2")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.3", label = TRUE) + ggtitle("res = 0.3")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.4", label = TRUE) + ggtitle("res = 0.4")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.5", label = TRUE) + ggtitle("res = 0.5")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.6", label = TRUE) + ggtitle("res = 0.6")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.7", label = TRUE) + ggtitle("res = 0.7")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.8", label = TRUE) + ggtitle("res = 0.8")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.0.9", label = TRUE) + ggtitle("res = 0.9")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.1", label = TRUE) + ggtitle("res = 1")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.1.1", label = TRUE) + ggtitle("res = 1.1")
DimPlot(obj_merged1, reduction = "umap", group.by = "SCT_snn_res.1.2", label = TRUE) + ggtitle("res = 1.2")

```
```{r}
Idents(obj_merged1) <- 'SCT_snn_res.1.2'
```

# highlight clusters

Cluster 0 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "0", highlight_color = "navy",background_color = "lightgray")
```

Cluster 1 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "1", highlight_color = "navy",background_color = "lightgray")
```
Cluster 2

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "2", highlight_color = "navy",background_color = "lightgray")
```

Cluster 3 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "3", highlight_color = "navy",background_color = "lightgray")
```
Cluster 4

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "4", highlight_color = "navy",background_color = "lightgray")
```
Cluster 5

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "5", highlight_color = "navy",background_color = "lightgray")
```

Cluster 6 -
Cluster 6
```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "6", highlight_color = "navy",background_color = "lightgray")
```
Cluster 7 


Cluster 7

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "7", highlight_color = "navy",background_color = "lightgray")
```

Cluster 8 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "8", highlight_color = "navy",background_color = "lightgray")

```

Cluster 9 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "9", highlight_color = "navy",background_color = "lightgray")

```
Cluster 10 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "10", highlight_color = "navy",background_color = "lightgray")

```
Cluster 11

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "11", highlight_color = "navy",background_color = "lightgray")
```
Cluster 12

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "12", highlight_color = "navy",background_color = "lightgray")

```

Cluster 13

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "13", highlight_color = "navy",background_color = "lightgray")

```

Cluster 14 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "14", highlight_color = "navy",background_color = "lightgray")

```

Cluster 15 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "15", highlight_color = "navy",background_color = "lightgray")

```
Cluster 16 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "16", highlight_color = "navy",background_color = "lightgray")
```

Cluster 17

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "17", highlight_color = "navy",background_color = "lightgray")

```
Cluster 18

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "18", highlight_color = "navy",background_color = "lightgray")
```
Cluster 19 

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "19", highlight_color = "navy",background_color = "lightgray")
```
Cluster 20

```{r}
Cluster_Highlight_Plot(seurat_object = obj_merged1, cluster_name = "20", highlight_color = "navy",background_color = "lightgray")
```
# Subclustering

Subclustering cluster 5

```{r}
subcluster <- FindSubCluster(obj_merged1, cluster = '5', graph.name = 'SCT_snn', resolution = 0.2, algorithm = 1)

Idents(subcluster) <- 'sub.cluster'
DimPlot(subcluster)

# 5_0 - this is neuroblast
Cluster_Highlight_Plot(seurat_object = subcluster, cluster_name = "5_0", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

# 5_1 - this is immature neurons 
Cluster_Highlight_Plot(seurat_object = subcluster, cluster_name = "5_1", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)
```
Subclustering cluster 6 

```{r}
subcluster <- FindSubCluster(subcluster, cluster = '6', graph.name = 'SCT_nn', resolution = 0.2, algorithm = 1)

Idents(subcluster) <- 'sub.cluster'
DimPlot(subcluster)

# 6_0 - 
Cluster_Highlight_Plot(seurat_object = subcluster, cluster_name = "6_0", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

# 6_1 -
Cluster_Highlight_Plot(seurat_object = subcluster, cluster_name = "6_1", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)
```

Subclustering cluster 9

```{r}
test1 <- subcluster
test1 <- FindSubCluster(test1, cluster = '9', graph.name = 'SCT_nn', resolution = 0.4, algorithm = 1)

Idents(test1) <- 'sub.cluster'
DimPlot(test1)

# 9_0 - 
Cluster_Highlight_Plot(seurat_object = test1, cluster_name = "9_0", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

# 9_1 -
Cluster_Highlight_Plot(seurat_object = test1, cluster_name = "9_1", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

Cluster_Highlight_Plot(seurat_object = test1, cluster_name = "9_2", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

Cluster_Highlight_Plot(seurat_object = test1, cluster_name = "9_3", highlight_color = "navy",background_color = "lightgray", split_seurat = TRUE)

DimPlot(test1)
```
# Renaming clusters

```{r}
test2 <- test1

test2 <- RenameIdents(object = test2,
"0" = "Astro-like 1",
"1" = "RGL-like 1",
"2" = "Astro-like 1",
"3" = "RGL-like 2",
"4" = "IFN-responsive glial/progenitor-like",
"5_0" = "Neuroblast-like",
"5_1" = "Immature neuron-like",
"6_0" = "Astro-like 1",
"6_1" = "Astro-like 2",
"7" = "IPC-like 1",
"8" = "Astro-like 1",
"9_0" = "RGL-like 1",
"9_1" = "Astro-like 1",
"9_2" = "IPC-like 1",
"9_3" = "RGL-like 1",
"10" = "IPC-like 4",
"11" = "Reactive astrocyte-like",
"12" = "IPC-like 2",
"13" = "Astro-like 3",
"14" = "IPC-like 2",
"15" = "Wound healing-like",
"16" = "IPC-like 3",
"17" = "IPC-like 3",
"18" = "Astro-like 1",
"19" = "Astro-like 4",
"20" = "Astro-like 1")

DimPlot(test2)
```

```{r}
seurat_pipeline_cluster <- test2
saveRDS(seurat_pipeline_cluster, file.path(RESULTS_DIR, "seurat_pipeline_cluster.rds"))
write.csv(Embeddings(seurat_pipeline_cluster, "umap"),file.path(RESULTS_DIR, "umap_coords_SCT_pca1-50_seed42.csv"))

```
