---
title: "03 Analysis"
output: html_document
---


```{r setup, include=FALSE}
# If running interactively from within /scripts, move up to project root
# If running interactively from within /scripts, move up to project root
if (basename(getwd()) == "scripts") setwd("..")

# Now we are in the project root
knitr::opts_knit$set(root.dir = normalizePath(".")) 
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)


source("scripts/00_setup.R")
```

# Load the seurat object

```{r}
# load diff
seurat_pipeline_cluster <- readRDS(file.path(RESULTS_DIR, "seurat_pipeline_cluster.rds"))

seurat_pipeline_cluster$active_ident <- Idents(seurat_pipeline_cluster)
```

# generating figure 2b

```{r}
new_order <- c("Astro-like 1", "Astro-like 2", "Astro-like 3", "Astro-like 4", "RGL-like 1", "RGL-like 2", "IPC-like 1", "IPC-like 2", "IPC-like 3", "IPC-like 4", "Neuroblast-like", "Immature neuron-like", "IFN-responsive glial/progenitor-like", "Reactive astrocyte-like", "Wound healing-like")

seurat_pipeline_cluster <- SetIdent(seurat_pipeline_cluster, value = factor(Idents(seurat_pipeline_cluster), levels = new_order))

colors_UMAP <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", 
  "#E55AA4", "#43BDB2", "#EECAA4", "#050B8D", "#FF6347",  
  "#9C73B4", "#B3D500", "#27601C", "#DC95E3", "#995D07")

UMAP_plot <- DimPlot(seurat_pipeline_cluster, cols = colors_UMAP) + labs(x = "UMAP1", y = "UMAP2")

ggsave(filename = "Fig2b.pdf", path =FIGURES_DIR , plot = UMAP_plot, width = 200, height = 100, device = cairo_pdf, units = "mm")
```

# generating figure 2c

Add timepoint to metadata:

```{r}
# Make a new metadata column
seurat_pipeline_cluster$timepoint <- 'Time'

# Fix meta data
metadata_seurat_pipeline_cluster <- seurat_pipeline_cluster@meta.data %>% 
  mutate(Time = case_when(
            sample.ID == "Proliferation control" ~ "Proliferation (2 days)",
            sample.ID == "Proliferation low TNF" ~ "Proliferation (2 days)",
            sample.ID == "Proliferation high TNF" ~ "Proliferation (2 days)",
            sample.ID == "Differentiation control (7 days)" ~ "Differentiation (7 days)",
            sample.ID == "Differentiation low TNF (7 days)" ~ "Differentiation (7 days)",
            sample.ID == "Differentiation high TNF (7 days)" ~ "Differentiation (7 days)",
            sample.ID == "Differentiation control (14 days)" ~ "Differentiation (14 days)",
            sample.ID == "Differentiation low TNF (14 days)" ~ "Differentiation (14 days)",
            sample.ID == "Differentiation high TNF (14 days)" ~ "Differentiation (14 days)"))

# Add new metadata
seurat_pipeline_cluster@meta.data <- metadata_seurat_pipeline_cluster
```
Add treatment column

```{r}
# Make a new metadata column
seurat_pipeline_cluster$treatment <- NULL

# Fix meta data
metadata_seurat_pipeline_cluster <- seurat_pipeline_cluster@meta.data %>% 
  mutate(treatment = case_when(
            sample.ID == "Proliferation control" ~ "Control",
            sample.ID == "Proliferation low TNF" ~ "Low TNF",
            sample.ID == "Proliferation high TNF" ~ "High TNF",
            sample.ID == "Differentiation control (7 days)" ~ "Control",
            sample.ID == "Differentiation low TNF (7 days)" ~ "Low TNF",
            sample.ID == "Differentiation high TNF (7 days)" ~ "High TNF",
            sample.ID == "Differentiation control (14 days)" ~ "Control",
            sample.ID == "Differentiation low TNF (14 days)" ~ "Low TNF",
            sample.ID == "Differentiation high TNF (14 days)" ~ "High TNF"))

# Add new metadata
seurat_pipeline_cluster@meta.data <- metadata_seurat_pipeline_cluster
```
Take a look at number of cells belonging to each cluster by line and timepoint. For this we make a contingency table.

```{r}
Prol <- subset(seurat_pipeline_cluster, Time =="Proliferation (2 days)")
Dif7 <- subset(seurat_pipeline_cluster,Time == "Differentiation (7 days)")
Dif14 <- subset(seurat_pipeline_cluster,Time == "Differentiation (14 days)")

Idents(Prol) <- "active_ident"
Idents(Dif7) <- "active_ident"
Idents(Dif14) <- "active_ident"
```

Look at cell frequencies in each sample:

```{r}
table(Idents(Prol), Prol$treatment) 
table(Idents(Dif7), Dif7$treatment)
table(Idents(Dif14), Dif14$treatment)

to_df <- function(df) {
  df %>% 
  as.data.frame() %>%
  as_tibble() %>% 
  dplyr::rename("active_ident"=Var1,"treatment"=Var2,"count"=Freq) %>% 
  arrange(active_ident)
}
```

Turn frequency tables into tidy data

```{r}
Prol <- table(Idents(Prol), Prol$treatment) %>%
  to_df() %>% 
  mutate(sample = "2-Day Proliferation") 

Dif7 <- table(Idents(Dif7), Dif7$treatment) %>% 
  to_df() %>% 
  mutate(sample = "7-Day Differentiation")

Dif14 <- table(Idents(Dif14), Dif14$treatment) %>% 
  to_df() %>% 
  mutate(sample = "14-Day Differentiation")

my_data <- Prol %>% 
  bind_rows(Dif7) %>% 
  bind_rows(Dif14) %>% 
  mutate(active_ident = factor(active_ident,
                            levels=c("Astro-like 1","Astro-like 2","Astro-like 3","Astro-like 4","RGL-like 1","RGL-like 2","IPC-like 1","IPC-like 2", "IPC-like 3","IPC-like 4", 'Neuroblast-like', 'Immature neuron-like', 'IFN-responsive glial/progenitor-like', 'Reactive astrocyte-like', 'Wound healing-like'))) %>% 
  mutate(sample = factor(sample,levels = c("2-Day Proliferation",
                                         "7-Day Differentiation",
                                         "14-Day Differentiation")))

```

Make stacked bar plots of cell type proportions

```{r}
my_data <- my_data %>%
  mutate(treatment = factor(treatment, levels = c("High TNF", "Low TNF", "Control")))

bar_plot <- my_data %>%
ggplot(aes(y=count,x=treatment,fill=active_ident)) + 
  geom_bar(position="fill",stat="identity") +
  coord_flip() +
  scale_fill_manual(values = colors_UMAP,name="") +
  scale_y_continuous(expand = c(0,0),
                     labels = percent_format()) +
  facet_wrap(~sample,ncol = 1) +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_text(size = 10, color = "black", face = "bold"),
        axis.ticks.y = element_blank(),
        strip.text = element_text(size = 10,face = "bold")) +
  labs(x="",y="Cluster Proportion") +
  theme(plot.margin = margin(0,1,0.5,0, "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(),plot.background = element_blank() )

ggsave(filename = "Fig2c.pdf", path =FIGURES_DIR , plot = bar_plot, width = 200, height = 120, device = cairo_pdf, units = "mm")

```

# generating figure 2d
```{r}
all_markers <- c('AQP4','S100B', 'GFAP', 'BCAN', 'CLU', 'ID4', 'FABP7', 'PTN', 'CDH4', 'CD44' ,'HES1','STMN1', 'MKI67', 'HMGB2','DCX','IGFBPL1' ,'PROX1','MAP2', 'NRXN3', 'TUBB3','SYNPR', 'DSCAM', 'CALB2')

dotplot <- DotPlot_scCustom(seurat_pipeline_cluster,features = all_markers,  x_lab_rotate = TRUE) 

ggsave(filename = "Fig2d.pdf", path =FIGURES_DIR , plot = dotplot, width = 240, height = 120, device = cairo_pdf, units = "mm")
```


#generating figure 2e
```{r}
human_genes <- read.csv("scripts/NIHMS1815814-supplement-1815814_Tab_4.csv")

top_twenty <- human_genes[1:20, ]
top_twenty_genes <- top_twenty$Gene

seurat_pipeline_cluster <- AddModuleScore(
  seurat_pipeline_cluster,
  features = list(top_twenty_genes),
  name = c('Immature DG cell signature'),
  search = TRUE,
  )

Immature_DG <- FeaturePlot_scCustom(
  seurat_pipeline_cluster,
  features     = "Immature DG cell signature1",          
  colors_use   = viridis_plasma_dark_high,    
  na_color     = "grey90",                    
  na_cutoff    = 0.15)    + xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() + ggtitle('Human imGCs signature')                         

ggsave(filename = "Fig2e.pdf", path =FIGURES_DIR , plot = Immature_DG, width = 200, height =200, device = cairo_pdf, units = "mm")

```


# generating figure 2f-g

```{r}
DefaultAssay(seurat_pipeline_cluster) <- "SCT"
seurat_pipeline_cluster <- PrepSCTFindMarkers(seurat_pipeline_cluster)
markers_clusters <-FindAllMarkers(seurat_pipeline_cluster,assay = 'SCT',  only.pos = TRUE, min.pct=0.25, test.use = "wilcox", logfc.threshold = 0.25)
markers_clusters %>% group_by(cluster) %>% arrange(desc(avg_log2FC)) %>% slice_head(n = 20) %>% ungroup() -> top20
```

```{r}
# get topp 100 genes from neuroblast and immature neuronal cluster 
neuroblast_top_100 <- markers_clusters %>% filter(cluster == 'Neuroblast-like') %>% arrange(desc(avg_log2FC)) %>% top_n(100, wt = avg_log2FC)
neuro_top_100 <- markers_clusters %>% filter(cluster == 'Immature neuron-like') %>% arrange(desc(avg_log2FC)) %>% top_n(100, wt = avg_log2FC)

# getting the genes
neuroblast_genes <- neuroblast_top_100$gene
neuro_genes <- neuro_top_100$gene

# convert gene symbols to entrez IDs 
entrez_ids_neuroblast <- bitr(neuroblast_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

entrez_ids_neuro<- bitr(neuro_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)
```

```{r}
# run GO enrinchment - BP
go_results_neuroblasts <- enrichGO(gene = entrez_ids_neuroblast$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

go_results_neuro <- enrichGO(gene = entrez_ids_neuro$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
```

Make dotplots 

```{r}
neuroblast_bb <- dotplot(go_results_neuroblasts, showCategory = 15)
ggsave(filename = "Fig2f.pdf", path = FIGURES_DIR , plot = neuroblast_bb, width = 200, height =200, device = cairo_pdf, units = "mm")
```

```{r}
neuro_bb <-dotplot(go_results_neuro, showCategory = 15)
ggsave(filename = "Fig2g.pdf", path = FIGURES_DIR , plot = neuro_bb, width = 200, height =200, device = cairo_pdf, units = "mm")
```

# generating figure 2h

```{r}
# Step 1: Subset to 7d and 14d differentiation samples
diff_cells <- subset(seurat_pipeline_cluster, subset = Time %in% c("Differentiation (7 days)", "Differentiation (14 days)"))

# Step 2: Subset only neuroblast and immature neurons
target_cells <- diff_cells@meta.data %>%filter(active_ident %in% c("Neuroblast-like", "Immature neuron-like"))

# Step 3: Count cells per treatment and type
counts <- target_cells %>%
  group_by(treatment, active_ident) %>%
  summarise(n = n(), .groups = "drop")

# Step 4: Get total number of cells per condition (in differentiation samples)
total_per_condition <- diff_cells@meta.data %>%
  group_by(treatment) %>%
  summarise(total = n(), .groups = "drop")

# Step 5: Merge counts and total, compute percentages
counts <- left_join(counts, total_per_condition, by = "treatment") %>%
  mutate(percent = (n / total) * 100)

counts$treatment <- factor(counts$treatment, levels = c("Control", "Low TNF", "High TNF"))

# Step 6: Plot
treatment_neurogenesis <- ggplot(counts, aes(x = treatment, y = percent, fill = active_ident)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Treatment", y = "Percentage of cells in neurogenic clusters", fill = "Cluster") +
  scale_fill_manual(values = c("Neuroblast-like" = "#9C73B4", "Immature neuron-like" = "#B3D500")) +
  theme_minimal() + theme(
    axis.text.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")) 

ggsave(filename = "Fig2h.pdf", path = FIGURES_DIR , plot = treatment_neurogenesis, width = 200, height =200, device = cairo_pdf, units = "mm")
```

# generate figure 3a

Generate heatmap comparing RGL vs. RGL immune 
```{r}
# Use then SCT assay - find markers expressed in minimum 25% of cells in the cluster
RGL_vs_RGL_immune <-FindMarkers(seurat_pipeline_cluster,ident.1 = 'RGL-like 2', ident.2 = 'RGL-like 1', assay = 'SCT',  only.pos = TRUE, min.pct=0.25, test.use = "wilcox", logfc.threshold = 0.25) %>% rownames_to_column(var = "gene")

# get the data from the top 40 genes in each cluster based on avg_log2FC
RGL_vs_RGL_immune %>% arrange(desc(avg_log2FC)) %>% slice_head(n = 40) %>% ungroup() -> top40_RGL_vs_RGL_immune

RGL_heatmap_genes <- top40_RGL_vs_RGL_immune$gene

# subset the seurat object 
RGL_all <- subset(seurat_pipeline_cluster, idents = c('RGL-like 2', 'RGL-like 1'))

# colors to use 
colors_RGL_heatmap <- c( "#0072B2", "#E55AA4")

# do heatmap - genes expressed in 25% of cells
heatmap_RGL_vs_RGL_immune <- DoHeatmap(RGL_all, features = RGL_heatmap_genes, assay = 'SCT', slot = "scale.data",label = FALSE, group.colors = colors_RGL_heatmap)

# save the plot 
ggsave(filename = "Fig3a.pdf", path = FIGURES_DIR , plot = heatmap_RGL_vs_RGL_immune, width = 200, height =200, device = cairo_pdf, units = "mm")
```
# generate figure 3b

```{r}
# Use then SCT assay - find markers expressed in minimum 25% of cells in the cluster
IPC_TNF_vs_IPCs <-FindMarkers(seurat_pipeline_cluster,ident.1 = 'IPC-like 4', ident.2 = c('IPC-like 1', 'IPC-like 2', 'IPC-like 3'), assay = 'SCT',  only.pos = TRUE, min.pct=0.25, test.use = "wilcox", logfc.threshold = 0.25) %>% rownames_to_column(var = "gene")

# get the data from the top 40 genes in each cluster based on avg_log2FC
IPC_TNF_vs_IPCs %>% arrange(desc(avg_log2FC)) %>% slice_head(n = 40) %>% ungroup() -> topIPC_TNF_vs_IPCs

IPC_heatmap_genes <- topIPC_TNF_vs_IPCs$gene

# subset the seurat object 
IPC_all <- subset(seurat_pipeline_cluster, idents = c('IPC-like 1', 'IPC-like 2', 'IPC-like 3', 'IPC-like 4'))

# colors to use 
colors_IPC_heatmap <- c("#43BDB2", "#EECAA4", "#050B8D", "#FF6347")

# do heatmap - genes expressed in 25% of cells
heatmap_IPC_vs_IPC_immune <- DoHeatmap(IPC_all, features = IPC_heatmap_genes, assay = 'SCT', slot = "scale.data",label = FALSE, group.colors = colors_IPC_heatmap)

# save the plot 
ggsave(filename = "Fig3b.pdf", path = FIGURES_DIR , plot = heatmap_IPC_vs_IPC_immune, width = 200, height =200, device = cairo_pdf, units = "mm")
```

# generate figure 3c

```{r}
IFN_genes <- c('MX1', 'ISG15', 'OAS1', 'OAS2', 'IFI6', 'IFI27', 'IFI35','IFI44', 'IFI44L','IFI16', 'IFIT1', 'IFIT3', 'STAT1', 'STAT2', 'TRIM22', 'BST2', 'CXCL10','IFITM1', 'IFITM3', 'B2M', 'IL32', 'HLA-B')

IFN_dotplot <-  DotPlot_scCustom(seurat_pipeline_cluster,features = IFN_genes,  x_lab_rotate = TRUE) 

ggsave(filename = "Fig3c.pdf", path = FIGURES_DIR , plot = IFN_dotplot, width = 325, height =125, device = cairo_pdf, units = "mm")
```

# generate figure d-f

```{r}
IFN_astroglial_top_100 <- markers_clusters %>% filter(cluster == 'IFN-responsive glial/progenitor-like') %>% arrange(desc(avg_log2FC)) %>% top_n(100, wt = avg_log2FC)

reactive_astro_top_100 <- markers_clusters %>% filter(cluster == 'Reactive astrocyte-like') %>% arrange(desc(avg_log2FC)) %>% top_n(100, wt = avg_log2FC)

wound_healing_top_100 <- markers_clusters %>% filter(cluster == 'Wound healing-like') %>% arrange(desc(avg_log2FC)) %>% top_n(100, wt = avg_log2FC)

# getting the genes
IFN_astroglial_top_100_genes <- IFN_astroglial_top_100$gene
reactive_astro_top_100_genes <- reactive_astro_top_100$gene
wound_healing_top_100_genes <- wound_healing_top_100$gene

# convert gene symbols to entrez IDs 
entrez_ids_IFN_astroglial_top_100_genes <- bitr(IFN_astroglial_top_100_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

entrez_ids_reactive_astro_top_100_genes<- bitr(reactive_astro_top_100_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

entrez_ids_wound_healing_top_100_genes<- bitr(wound_healing_top_100_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)
```


Go enrichment - BP - biologial process 
```{r}
# run GO enrinchment - BP
go_results_IFN_astroglial <- enrichGO(gene = entrez_ids_IFN_astroglial_top_100_genes$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

go_results_reactive_astro <- enrichGO(gene = entrez_ids_reactive_astro_top_100_genes$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

go_results_wound_healing <- enrichGO(gene = entrez_ids_wound_healing_top_100_genes$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
```
Dot plots:

```{r}
go_results_IFN_astroglial_bb <- dotplot(go_results_IFN_astroglial, showCategory = 15) 
ggsave(filename = "Fig3d.pdf", path = FIGURES_DIR , plot = go_results_IFN_astroglial_bb, width = 150, height =250, device = cairo_pdf, units = "mm")
```

```{r}
go_results_reactive_astro_bb <- dotplot(go_results_reactive_astro, showCategory = 15) 
ggsave(filename = "Fig3e.pdf", path = FIGURES_DIR , plot = go_results_reactive_astro_bb, width = 150, height =300, device = cairo_pdf, units = "mm")
```

```{r}
go_results_wound_healing_bb <- dotplot(go_results_wound_healing, showCategory = 15) 
ggsave(filename = "Fig3f.pdf", path = FIGURES_DIR , plot = go_results_wound_healing_bb, width = 150, height =200, device = cairo_pdf, units = "mm")
```

# generate figure 3g

IFN-beta NSC signature - https://www.embopress.org/doi/full/10.15252/emmm.202216434 

```{r}
IFN_B_response_genes_df <- read.csv("scripts/NSC_IFN_b_genes.csv")
IFN_B_response_genes <- IFN_B_response_genes_df$Genes

seurat_pipeline_cluster <- AddModuleScore(
  seurat_pipeline_cluster,
  features = list(IFN_B_response_genes),
  name = c('NSC IFN-β signature'),
  search = TRUE,
)

IFN_signature <- FeaturePlot_scCustom(
  seurat_pipeline_cluster,
  features     = "NSC IFN-β signature1",            # your module score metadata
  colors_use   = viridis_plasma_dark_high,     # or your preferred gradient
  na_color     = "grey90",                     # color for low module scores
  na_cutoff    = 0.05)    + xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() + ggtitle('NSC IFN-β signature')                         # everything ≤ 0 is grey; >0 uses gradient

ggsave(filename = "Fig3g.pdf", path = FIGURES_DIR , plot = IFN_signature, width = 325, height =125, device = cairo_pdf, units = "mm")
```



# generate figure 3h

```{r}
# generate net 
net <- get_collectri(organism='human', split_complexes=FALSE)

# Extract the normalized log-transformed counts
mat <- seurat_pipeline_cluster[["SCT"]]$data

# Run ulm
acts <- run_ulm(mat=mat, net=net, .source='source', .target='target',.mor='mor', minsize = 5)
```


```{r}
# Extract ulm and store it in tfsulm 
seurat_pipeline_cluster[['tfsulm']] <- acts %>%
  pivot_wider(id_cols = 'source', names_from = 'condition',
              values_from = 'score') %>%
  column_to_rownames('source') %>%
  Seurat::CreateAssayObject(.)

# Change assay
DefaultAssay(object = seurat_pipeline_cluster) <- "tfsulm"

# Scale the data
seurat_pipeline_cluster <- ScaleData(seurat_pipeline_cluster)
seurat_pipeline_cluster@assays$tfsulm@data <- seurat_pipeline_cluster@assays$tfsulm@scale.data
```


```{r}
n_tfs <- 50

Idents(seurat_pipeline_cluster) <- seurat_pipeline_cluster$active_ident

# Extract activities from object as a long dataframe
df <- t(as.matrix(seurat_pipeline_cluster@assays$tfsulm$data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(seurat_pipeline_cluster)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Get top tfs with more variable means across clusters
tfs <- df %>%
  group_by(source) %>%
  summarise(std = sd(mean)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)

# Subset long data frame to top tfs and transform to wide matrix
top_acts_mat <- df %>%
  filter(source %in% tfs) %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()

# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-3, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 3, length.out=floor(palette_length/2)))

mat2 <- t(top_acts_mat)

# Plot
pheatmap(mat2,border_color = NA, color=my_color, breaks = my_breaks, filename = 'figures/fig3h.pdf',width  = 7,height = 10)
```

# extended data fig 1a

```{r}
DefaultAssay(object = seurat_pipeline_cluster) <- "SCT"

Idents(seurat_pipeline_cluster) <- 'sample.ID'

seurat_pipeline_cluster$sample.ID <- factor(seurat_pipeline_cluster$sample.ID , levels = c("Proliferation control","Proliferation low TNF","Proliferation high TNF","Differentiation control (7 days)","Differentiation low TNF (7 days)","Differentiation high TNF (7 days)","Differentiation control (14 days)","Differentiation low TNF (14 days)","Differentiation high TNF (14 days)"))


Idents(seurat_pipeline_cluster) <- 'active_ident'

new_order <- c("Astro-like 1", "Astro-like 2", "Astro-like 3", "Astro-like 4", "RGL-like 1", "RGL-like 2", "IPC-like 1", "IPC-like 2", "IPC-like 3", "IPC-like 4", "Neuroblast-like", "Immature neuron-like", "IFN-responsive glial/progenitor-like", "Reactive astrocyte-like", "Wound healing-like")

seurat_pipeline_cluster <- SetIdent(seurat_pipeline_cluster, value = factor(Idents(seurat_pipeline_cluster), levels = new_order))

sample_ID_plot <- DimPlot_scCustom(
  seurat_pipeline_cluster,
  label = FALSE,
  split.by = "sample.ID",
  repel = TRUE,
  num_columns = 3,
  split_seurat = TRUE,
  colors_use = colors_UMAP,
  pt.size = 0.1
) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  NoAxes() +
  theme(
    strip.text = element_text(size =9),  # 👈 changes panel title size
    legend.title = element_text(size = 5))

ggsave(filename = "extended_fig1a.pdf", path = FIGURES_DIR , plot = sample_ID_plot, width = 325, height =125, device = cairo_pdf, units = "mm")
```

# extended data fig 1b

```{r}
# subset prol data
prol <- subset(seurat_pipeline_cluster, subset = Time == 'Proliferation (2 days)')

prol$treatment <- factor(x = prol$treatment, levels = c('Control','Low TNF','High TNF'))

Idents(prol) <- 'treatment'

# violin plots 
ISG15 <- VlnPlot(prol, features = 'ISG15')
IFI27 <- VlnPlot(prol, features = 'IFI27')
IFI6 <- VlnPlot(prol, features = 'IFI6')
IFITM3 <- VlnPlot(prol, features = 'IFITM3')

dose_dependent1 <- ISG15 |IFI27
dose_dependent2 <- IFI6 |IFITM3

ggsave(filename = "extended_fig1b1.pdf", path = FIGURES_DIR , plot = dose_dependent1, width = 325, height =125, device = cairo_pdf, units = "mm")
ggsave(filename = "extended_fig1b2.pdf", path = FIGURES_DIR , plot = dose_dependent2, width = 325, height =125, device = cairo_pdf, units = "mm")
```

# extended data fig 1c

```{r}
RGLs_markers <-FindMarkers(seurat_pipeline_cluster,ident.1 = 'RGL-like 2', ident.2 = 'RGL-like 1', assay = 'SCT',  only.pos = TRUE, min.pct=0.25, test.use = "wilcox", logfc.threshold = 0.25) %>% rownames_to_column(var = "gene")

RGLs_markers %>% arrange(desc(avg_log2FC)) %>% slice_head(n = 100) %>% ungroup() -> RGL_results

RGL_results_genes <- RGL_results$gene

# convert gene symbols to entrez IDs 
RGL_entrez <- bitr(RGL_results_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

RGL_BP <- enrichGO(gene = RGL_entrez$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

RGL_downregulated_tnf <- dotplot(RGL_BP, showCategory = 15)

ggsave(filename = "extended_fig1c.pdf", path = FIGURES_DIR , plot = RGL_downregulated_tnf, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# extended data fig 1d

```{r}
IPC_down_TNF <-FindMarkers(seurat_pipeline_cluster,ident.1 = 'IPC-like 4', ident.2 = c('IPC-like 1', 'IPC-like 2', 'IPC-like 3'), assay = 'SCT',  only.pos = TRUE, min.pct=0.25, test.use = "wilcox", logfc.threshold = 0.25) %>% rownames_to_column(var = "gene")

IPC_down_TNF %>% arrange(desc(avg_log2FC)) %>% slice_head(n = 100) %>% ungroup() -> IPC_down_TNF_top_100

IPC_down_TNF_top_100_genes <- IPC_down_TNF_top_100$gene

IPC_TNF_down <- bitr(IPC_down_TNF_top_100_genes, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

IPC_TNF_down_BP <- enrichGO(gene = IPC_TNF_down$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "BP",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

extended_fig1d <- dotplot(IPC_TNF_down_BP, showCategory = 15)

ggsave(filename = "extended_fig1d.pdf", path = FIGURES_DIR , plot = extended_fig1d, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# extended data fig 1e-f

```{r}
DefaultAssay(object = seurat_pipeline_cluster) <- "tfsulm"

STAT1 <- FeaturePlot_scCustom(
  seurat_pipeline_cluster,
  features     = "STAT1",            # your module score metadata
  colors_use   = viridis_plasma_dark_high,     # or your preferred gradient
  na_color     = "grey90",                     # color for low module scores
  na_cutoff    = 0.5)    + xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() + ggtitle('STAT1 activity')                         # everything ≤ 0 is grey; >0 uses gradient

ggsave(filename = "extended_fig1e.pdf", path = FIGURES_DIR , plot = STAT1, width = 200, height =150, device = cairo_pdf, units = "mm")

STAT2 <- FeaturePlot_scCustom(
  seurat_pipeline_cluster,
  features     = "STAT2",            # your module score metadata
  colors_use   = viridis_plasma_dark_high,     # or your preferred gradient
  na_color     = "grey90",                     # color for low module scores
  na_cutoff    = 0.5)    + xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() + ggtitle('STAT2 activity')                         # everything ≤ 0 is grey; >0 uses gradient

ggsave(filename = "extended_fig1f.pdf", path = FIGURES_DIR , plot = STAT2, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 8

```{r}
go_results_neuro_cc <- enrichGO(gene = entrez_ids_neuro$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "CC",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

cc_neuro_dotplot <- dotplot(go_results_neuro_cc, showCategory = 15) 

ggsave(filename = "supplementary_fig8.pdf", path = FIGURES_DIR , plot = cc_neuro_dotplot, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9a

```{r}
DefaultAssay(seurat_pipeline_cluster) <- 'SCT'

# Parameters - change gene and object name if needed
gene <- "TNFRSF1A"   # <- replace with your gene name
seu_obj <- seurat_pipeline_cluster             

# Check gene presence
if (!(gene %in% rownames(seu_obj))) {
  stop(paste0("Gene '", gene, "' not found in the Seurat object."))
}

# Pull expression data
expr_mat <- GetAssayData(seu_obj, assay = DefaultAssay(seu_obj), slot = "data")
expr_vec <- as.numeric(expr_mat[gene, ])

# Build dataframe with cluster identity and expression
df <- data.frame(
  cluster = as.character(Idents(seu_obj)),
  expr = expr_vec,
  stringsAsFactors = FALSE
)

# Convert cluster to numeric if possible (to ensure natural order)
df <- df %>%
  mutate(cluster_num = suppressWarnings(as.numeric(cluster))) %>%
  mutate(cluster_num = ifelse(is.na(cluster_num), cluster, cluster))  # fallback for non-numeric

# Summarise: percent of cells with expr > 0 per cluster
summary_df <- df %>%
  group_by(cluster, cluster_num) %>%
  summarise(
    n_cells = n(),
    n_expressing = sum(expr > 0),
    pct_expressing = 100 * (n_expressing / n_cells),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(cluster_num))

# Ensure numeric ordering of x-axis
summary_df$cluster <- factor(summary_df$cluster, levels = summary_df$cluster)

# Plot
p1 <- ggplot(summary_df, aes(x = cluster, y = pct_expressing)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = paste0(round(pct_expressing, 1), "%")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), limits = c(0, NA)) +
  labs(
    title = paste0("Percent of cells expressing ", gene, " per cluster"),
    x = "Cluster",
    y = "Percent expressing (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

ggsave(filename = "supplementary_fig9a.pdf", path = FIGURES_DIR , plot = p1, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9b

```{r}
DefaultAssay(seurat_pipeline_cluster) <- 'SCT'

# Parameters - change gene and object name if needed
gene <- "TNFRSF1B"   # <- replace with your gene name
seu_obj <- seurat_pipeline_cluster             

# Check gene presence
if (!(gene %in% rownames(seu_obj))) {
  stop(paste0("Gene '", gene, "' not found in the Seurat object."))
}

# Pull expression data
expr_mat <- GetAssayData(seu_obj, assay = DefaultAssay(seu_obj), slot = "data")
expr_vec <- as.numeric(expr_mat[gene, ])

# Build dataframe with cluster identity and expression
df <- data.frame(
  cluster = as.character(Idents(seu_obj)),
  expr = expr_vec,
  stringsAsFactors = FALSE
)

# Convert cluster to numeric if possible (to ensure natural order)
df <- df %>%
  mutate(cluster_num = suppressWarnings(as.numeric(cluster))) %>%
  mutate(cluster_num = ifelse(is.na(cluster_num), cluster, cluster))  # fallback for non-numeric

# Summarise: percent of cells with expr > 0 per cluster
summary_df <- df %>%
  group_by(cluster, cluster_num) %>%
  summarise(
    n_cells = n(),
    n_expressing = sum(expr > 0),
    pct_expressing = 100 * (n_expressing / n_cells),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(cluster_num))

# Ensure numeric ordering of x-axis
summary_df$cluster <- factor(summary_df$cluster, levels = summary_df$cluster)

# Plot
p2 <- ggplot(summary_df, aes(x = cluster, y = pct_expressing)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = paste0(round(pct_expressing, 1), "%")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), limits = c(0, NA)) +
  labs(
    title = paste0("Percent of cells expressing ", gene, " per cluster"),
    x = "Cluster",
    y = "Percent expressing (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
ggsave(filename = "supplementary_fig9b.pdf", path = FIGURES_DIR , plot = p2, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9c

```{r}
DefaultAssay(seurat_pipeline_cluster) <- 'SCT'
TNFR1_violin <- VlnPlot_scCustom(seurat_pipeline_cluster,features     = "TNFRSF1A", colors_use = colors_UMAP)
ggsave(filename = "supplementary_fig9c.pdf", path = FIGURES_DIR , plot = TNFR1_violin, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9d
```{r}
DefaultAssay(seurat_pipeline_cluster) <- 'SCT'
TNFR2_violin <- VlnPlot_scCustom(seurat_pipeline_cluster,features     = "TNFRSF1B", colors_use = colors_UMAP)
ggsave(filename = "supplementary_fig9d.pdf", path = FIGURES_DIR , plot = TNFR2_violin, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9e

```{r}
TNFRSF1A_dotplot <- DotPlot(seurat_pipeline_cluster,features     = c("TNFRSF1A"))
ggsave(filename = "supplementary_fig9e.pdf", path = FIGURES_DIR , plot = TNFRSF1A_dotplot, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 9f

```{r}
TNFRSF1B_dotplot <-DotPlot(seurat_pipeline_cluster,features     = c("TNFRSF1B"))
ggsave(filename = "supplementary_fig9f.pdf", path = FIGURES_DIR , plot = TNFRSF1B_dotplot, width = 200, height =150, device = cairo_pdf, units = "mm")

```

# supplementary Fig. 10a and c

```{r}
ensembl <- useMart("ensembl",dataset="hsapiens_gene_ensembl")

apoptosis <- getBM(attributes=c('hgnc_symbol', 'reactome'),
                   filters = 'reactome', 
                   values = 'R-HSA-109581', mart = ensembl)

apoptosis_genes <- unique(apoptosis$hgnc_symbol)
```

```{r}
seurat_pipeline_cluster <- AddModuleScore(
  seurat_pipeline_cluster,
  features = list(apoptosis_genes),
  pool = NULL,
  nbin = 24,
  ctrl = 100,
  k = FALSE,
  assay = NULL,
  name = c("apoptosis_score"),
  seed = 1,
  search = TRUE,
)

apoptosis_Violinplot <- VlnPlot(seurat_pipeline_cluster,features = "apoptosis_score1")+ labs(title = "Apoptosis (R-HSA-109581)")

ggsave(filename = "supplementary_fig10a.pdf", path = FIGURES_DIR , plot = apoptosis_Violinplot, width = 200, height =150, device = cairo_pdf, units = "mm")
```

```{r}
seurat_pipeline_cluster$treatment <- factor(seurat_pipeline_cluster$treatment,
                          levels = c("Control", "Low TNF", "High TNF"))

apoptosis_score_1 <- VlnPlot(seurat_pipeline_cluster, features = "apoptosis_score1", group.by = "treatment") + 
labs(title = "Apoptosis") + theme_minimal(base_size = 14)

ggsave(filename = "supplementary_fig10c.pdf", path = FIGURES_DIR , plot = apoptosis_score_1, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 10b and d

```{r}
ensembl <- useMart("ensembl",dataset="hsapiens_gene_ensembl")

apoptosis_2 <- getBM(attributes=c('hgnc_symbol', 'reactome'),
                   filters = 'reactome', 
                   values = 'R-HSA-5357801', mart = ensembl)

apoptosis_genes_2 <- unique(apoptosis_2$hgnc_symbol)
```

```{r}
seurat_pipeline_cluster <- AddModuleScore(
  seurat_pipeline_cluster,
  features = list(apoptosis_genes_2),
  pool = NULL,
  nbin = 24,
  ctrl = 100,
  k = FALSE,
  assay = NULL,
  name = c("apoptosis_score_2"),
  seed = 1,
  search = TRUE,
)
```

```{r}
programmed_cell_death_Violinplot_2 <- VlnPlot(seurat_pipeline_cluster,features = "apoptosis_score_21")+ labs(title = "Programmed Cell Death (R-HSA-5357801)")

ggsave(filename = "supplementary_fig10b.pdf", path = FIGURES_DIR , plot = programmed_cell_death_Violinplot_2, width = 200, height =150, device = cairo_pdf, units = "mm")

apoptosis_score_2 <- VlnPlot(seurat_pipeline_cluster, features = "apoptosis_score_21", group.by = "treatment") + labs(title = "Programmed Cell Death") +theme_minimal(base_size = 14)

ggsave(filename = "supplementary_fig10d.pdf", path = FIGURES_DIR , plot = apoptosis_score_2, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 11

RGL
```{r}
RGL_CC <- enrichGO(gene = RGL_entrez$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "CC",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

RGL_CC_dotplot <- dotplot(RGL_CC, showCategory = 15)

ggsave(filename = "supplementary_fig11_RGL.pdf", path = FIGURES_DIR , plot = RGL_CC_dotplot, width = 200, height =150, device = cairo_pdf, units = "mm")
```

IPC:

```{r}
IPC_CC <- enrichGO(gene = IPC_TNF_down$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       keyType = "ENTREZID",
                       ont = "CC",       # Can be "BP", "MF", or "CC"
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)

IPC_CC_dotplot <- dotplot(IPC_CC, showCategory = 15)

ggsave(filename = "supplementary_fig11_IPC.pdf", path = FIGURES_DIR , plot = IPC_CC_dotplot, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 12

```{r}
IFNAR1 <- FeaturePlot(seurat_pipeline_cluster, features = "IFNAR1", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig12_ifnar1.pdf", path = FIGURES_DIR , plot = IFNAR1, width = 200, height =150, device = cairo_pdf, units = "mm")

IFNAR2 <- FeaturePlot(seurat_pipeline_cluster, features = "IFNAR2", order = TRUE)+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig12_ifnar2.pdf", path = FIGURES_DIR , plot = IFNAR2, width = 200, height =150, device = cairo_pdf, units = "mm")

JAK1 <- FeaturePlot(seurat_pipeline_cluster, features = "JAK1", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig12_jak1.pdf", path = FIGURES_DIR , plot = JAK1, width = 200, height =150, device = cairo_pdf, units = "mm")

TYK2 <- FeaturePlot(seurat_pipeline_cluster, features = "TYK2", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig12_tyk2.pdf", path = FIGURES_DIR , plot = TYK2, width = 200, height =150, device = cairo_pdf, units = "mm")

STAT1 <- FeaturePlot(seurat_pipeline_cluster, features = "STAT1", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed()

ggsave(filename = "supplementary_fig12_stat1.pdf", path = FIGURES_DIR , plot = STAT1, width = 200, height =150, device = cairo_pdf, units = "mm")

STAT2 <- FeaturePlot(seurat_pipeline_cluster, features = "STAT2", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig12_stat2.pdf", path = FIGURES_DIR , plot = STAT2, width = 200, height =150, device = cairo_pdf, units = "mm")

IRF9 <- FeaturePlot(seurat_pipeline_cluster, features = "IRF9", order = TRUE )+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed()

ggsave(filename = "supplementary_fig12_irf9.pdf", path = FIGURES_DIR , plot = IRF9, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 14a

```{r}

CGAS <- FeaturePlot(seurat_pipeline_cluster, features = "CGAS", order = TRUE)+ xlab('UMAP1') +  ylab('UMAP2') + theme(plot.title = element_text(hjust = 0.5)) + coord_fixed() 

ggsave(filename = "supplementary_fig14.pdf", path = FIGURES_DIR , plot = CGAS, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 19a

```{r}
# Parameters - change gene and object name if needed
gene <- "CXCL10"   
seu_obj <- seurat_pipeline_cluster             

# Check gene presence
if (!(gene %in% rownames(seu_obj))) {
  stop(paste0("Gene '", gene, "' not found in the Seurat object."))
}

# Pull expression data
expr_mat <- GetAssayData(seu_obj, assay = DefaultAssay(seu_obj), slot = "data")
expr_vec <- as.numeric(expr_mat[gene, ])

# Build dataframe with cluster identity and expression
df <- data.frame(
  cluster = as.character(Idents(seu_obj)),
  expr = expr_vec,
  stringsAsFactors = FALSE
)

# Convert cluster to numeric if possible (to ensure natural order)
df <- df %>%
  mutate(cluster_num = suppressWarnings(as.numeric(cluster))) %>%
  mutate(cluster_num = ifelse(is.na(cluster_num), cluster, cluster))  # fallback for non-numeric

# Summarise: percent of cells with expr > 0 per cluster
summary_df <- df %>%
  group_by(cluster, cluster_num) %>%
  summarise(
    n_cells = n(),
    n_expressing = sum(expr > 0),
    pct_expressing = 100 * (n_expressing / n_cells),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(cluster_num))

# Ensure numeric ordering of x-axis
summary_df$cluster <- factor(summary_df$cluster, levels = summary_df$cluster)

# Plot
CXCL10 <- ggplot(summary_df, aes(x = cluster, y = pct_expressing)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = paste0(round(pct_expressing, 1), "%")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), limits = c(0, NA)) +
  labs(
    title = paste0("Percent of cells expressing ", gene, " per cluster"),
    x = "Cluster",
    y = "Percent expressing (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

ggsave(filename = "supplementary_fig19a.pdf", path = FIGURES_DIR , plot = CXCL10, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 19b

```{r}
# Parameters - change gene and object name if needed
gene <- "CXCL11"   
seu_obj <- seurat_pipeline_cluster             

# Check gene presence
if (!(gene %in% rownames(seu_obj))) {
  stop(paste0("Gene '", gene, "' not found in the Seurat object."))
}

# Pull expression data
expr_mat <- GetAssayData(seu_obj, assay = DefaultAssay(seu_obj), slot = "data")
expr_vec <- as.numeric(expr_mat[gene, ])

# Build dataframe with cluster identity and expression
df <- data.frame(
  cluster = as.character(Idents(seu_obj)),
  expr = expr_vec,
  stringsAsFactors = FALSE
)

# Convert cluster to numeric if possible (to ensure natural order)
df <- df %>%
  mutate(cluster_num = suppressWarnings(as.numeric(cluster))) %>%
  mutate(cluster_num = ifelse(is.na(cluster_num), cluster, cluster))  # fallback for non-numeric

# Summarise: percent of cells with expr > 0 per cluster
summary_df <- df %>%
  group_by(cluster, cluster_num) %>%
  summarise(
    n_cells = n(),
    n_expressing = sum(expr > 0),
    pct_expressing = 100 * (n_expressing / n_cells),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(cluster_num))

# Ensure numeric ordering of x-axis
summary_df$cluster <- factor(summary_df$cluster, levels = summary_df$cluster)

# Plot
CXCL11 <- ggplot(summary_df, aes(x = cluster, y = pct_expressing)) +
  geom_col(width = 0.7, fill = "#4682B4") +
  geom_text(aes(label = paste0(round(pct_expressing, 1), "%")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), limits = c(0, NA)) +
  labs(
    title = paste0("Percent of cells expressing ", gene, " per cluster"),
    x = "Cluster",
    y = "Percent expressing (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

ggsave(filename = "supplementary_fig19b.pdf", path = FIGURES_DIR , plot = CXCL11, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 19c

```{r}
CXCL10_violin <- VlnPlot_scCustom(seurat_pipeline_cluster,features     = "CXCL10", colors_use = colors_UMAP)

ggsave(filename = "supplementary_fig19c.pdf", path = FIGURES_DIR , plot = CXCL10_violin, width = 200, height =150, device = cairo_pdf, units = "mm")
```

# supplementary Fig. 19d

```{r}
CXCL11_violin <- VlnPlot_scCustom(seurat_pipeline_cluster,features     = "CXCL11", colors_use = colors_UMAP)

ggsave(filename = "supplementary_fig19d.pdf", path = FIGURES_DIR , plot = CXCL11_violin, width = 200, height =150, device = cairo_pdf, units = "mm")
```




